Probing heavy ion collisions using quark and gluon jet substructure

### 2018-03-06 qg_ML.tex
text and ref additions

The use of quark jets and gluon jets as independent hard probes of heavy ion collisions is closely related to the tagging of these two probes, which is an outstanding problem \cite{Gras:2017jty,Frye:2017yrw}. The lowest-order feature that separates quark jets from gluon jets is the different color charges carried by the jet-initiating partons. The Casimir factors of quark jets and gluon jets are $C_F = 4/3$ and $C_A=3$, respectively, and the larger color charge of gluons is expected to result in broader spread of the radiation inside such jets. Additional distinctive features can be included to improve the tagging by combining a variety of jet substructure variables in a multivariate analysis, similar to how heavy flavor jets are identified at the LHC \cite{}. Driven by the need to identify and study boosted objects at the LHC, the high energy community has made significant advances on the use of machine-learning techniques leading to higher tagging efficiency in a wide kinematic range \cite{Komiske:2016rsd, Larkoski:2017jix}.

@@

To use quark and gluon jets as independent probes of heavy ion collisions, we need to first define and identify the respective probes. In this paper, we define jet flavor based on the parton from hard scattering that initiates the parton shower evolution. We use the \jw Monte Carlo simulation, a perturbative jet evolution framework to generate proton-proton and central (0-20\%) lead-lead collisions at $\sqrt{s}=2.76$ TeV using the standard setup \cite{Zapp:2013zya}. We use the prompt photon production channels \cite{KunnawalkamElayavalli:2016ttl} $q +\gamma$ and $g +\gamma$ to generate quark and gluon jet samples, respectively, as an attempt to prepare two sets of jets with distinct quark and gluon jet fractions. It is possible to construct jet samples by performing hard process and kinematical selections (such as dijet v.s. $Z$+jet) to enhance either the gluon or the quark jet fraction. On the other hand, weak and unsupervised learning algorithms \cite{Metodiev:2017vrx,Komiske:2018oaa,Metodiev:2018ftz} can also be employed in the near future on such samples in order to treat both data and simulations on equal footing.

@@

\footnote{We thank Patrick Komiske and Eric Metodiev for making the beautiful, ``{\sl Ripples in Jets}" illustration in Figure~\ref{fig:tjetQCD} and the one in the Appendix, and we congratulate again the Nobel prize award to gravitational wave detection.} The multivariate analysis and jet image recognition discussed previously are two characteristically different methods. The former uses physics-motivated observables and is a ``bottom-up" approach to collect useful jet features. However, the set of observables identified may be highly correlated, and it is not clear how to systematically improve the method. On the other hand, jet images represent low-level, comprehensive jet information. The training of a deep neural network is a ``top-down" approach to identify all useful features through model optimization. However, extracting physical messages from the trained neural network parameters can be a challenging task.

We develop and apply the telescoping deconstruction (TD) framework in our study of quark and gluon jet substructures in pp and PbPb collisions. TD aims to embrace both advantages of multivariate analysis and jet image recognition. It decomposes jets using physics-motivated, comprehensive sets of observables which form an organized fragmentation basis of jet substructures. It is a complete and systematic subjet expansion consisting simply of subjet kinematic variables, i.e. the subjet $p_T$ and mass. The expansion is ordered by $N$, the number of subjets exclusively reconstructed, and the individual TD observables are physically motivated to encode the hard splitting of jets and relevant non-perturbative physics within each subjet variable. The telescoping procedure probes radiation around dominant energy flow in a jet with multiple angular resolutions \cite{Chien:2013kca,Adams:2015hiv,Chien:2014hla,Chien:2017xrb}. It efficiently quantifies the radiation distribution by first capturing the dominant energy in the subjet reconstruction followed by reaching out to include wide-angle, soft radiation. Figure~\ref{fig:tjetQCD} illustrates the telescoping deconstruction of a random pp jet at T1, T2 and T3 orders (clockwise from top left). Details about the telescoping deconstruction procedures are provided in the Appendix, including the applications to boosted $W$ and top tagging in high energy physics.

>>>>

The use of quark jets and gluon jets as independent hard probes of heavy ion collisions is closely related to the tagging of these two probes, which is an outstanding problem \cite{Gras:2017jty,Frye:2017yrw}. The lowest-order feature that separates quark jets from gluon jets is the different color charges carried by the jet-initiating partons. The Casimir factors of quark jets and gluon jets are $C_F = 4/3$ and $C_A=3$, respectively, and the larger color charge of gluons is expected to result in broader spread of the radiation inside such jets. Additional distinctive features can be included to improve the tagging by combining a variety of jet substructure variables in a multivariate analysis, similar to how heavy flavor jets are identified at the LHC \cite{Chatrchyan:2012jua, Aad:2015ydr}. Driven by the need to identify and study boosted objects at the LHC, the high energy community has made significant advances on the use of machine-learning techniques leading to higher tagging efficiency in a wide kinematic range \cite{Komiske:2016rsd, Larkoski:2017jix}.

@@

To use quark and gluon jets as independent probes of heavy ion collisions, we need to first define and identify the respective probes. In this paper, we define jet flavor based on the parton from hard scattering that initiates the parton shower evolution. We use the \jw Monte Carlo simulation, a perturbative jet evolution framework to generate proton-proton and central (0-20\%) lead-lead collisions at $\sqrt{s}=2.76$ TeV using the standard setup \cite{Zapp:2013zya}. We use the prompt photon production channels \cite{KunnawalkamElayavalli:2016ttl} $q +\gamma$ and $g +\gamma$ to generate quark and gluon jet samples, respectively, as an attempt to prepare two sets of jets with distinct quark and gluon jet fractions. It is possible to construct jet samples by performing hard process and kinematical selections (such as dijet v.s. $Z$+jet) to enhance either the gluon or the quark jet fraction. On the other hand, weak and unsupervised learning algorithms \cite{Dery:2017fap, Cohen:2017exh, Metodiev:2017vrx,Komiske:2018oaa,Metodiev:2018ftz} can also be employed in the near future on such samples in order to treat both data and simulations on equal footing.

@@

The multivariate analysis and jet image recognition discussed previously are two characteristically different methods. The former uses physics-motivated observables and is a ``bottom-up" approach to collect useful jet features. However, the set of observables identified may be highly correlated, and it is not clear how to systematically improve the method. On the other hand, jet images represent low-level, comprehensive jet information. The training of a deep neural network is a ``top-down" approach to identify all useful features through model optimization. However, extracting physical messages from the trained neural network parameters can be a challenging task.

We develop and apply the telescoping deconstruction (TD) framework in our study of quark and gluon jet substructures in pp and PbPb collisions. TD aims to embrace both advantages of multivariate analysis and jet image recognition. It decomposes jets using physics-motivated, comprehensive sets of observables which form an organized fragmentation basis of jet substructures. It is a complete and systematic subjet expansion~\footnote{Orthogonal frameworks utilizing jet information have been published recently~\cite{Datta:2017rhs, Komiske:2017aww}} consisting simply of subjet kinematic variables, i.e. the subjet $p_T$ and mass. The expansion is ordered by $N$, the number of subjets exclusively reconstructed, and the individual TD observables are physically motivated to encode the hard splitting of jets and relevant non-perturbative physics within each subjet variable. The telescoping procedure probes radiation around dominant energy flow in a jet with multiple angular resolutions \cite{Chien:2013kca,Adams:2015hiv,Chien:2014hla,Chien:2017xrb}. It efficiently quantifies the radiation distribution by first capturing the dominant energy in the subjet reconstruction followed by reaching out to include wide-angle, soft radiation. Figure~\ref{fig:tjetQCD}\footnote{We thank Patrick Komiske and Eric Metodiev for making the beautiful, ``{\sl Ripples in Jets}" illustration in Figure~\ref{fig:tjetQCD} and the one in the Appendix, inspired by the nobel prize awarded to the detection of gravitational waves i.e. {\sl Ripples in Space-Time}.} illustrates the telescoping deconstruction of a random pp jet at T1, T2 and T3 orders (clockwise from top left). Details about the telescoping deconstruction procedures are provided in the Appendix, including the applications to boosted $W$ and top tagging in high energy physics.
